COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-08-22 TACO-paper-v2/coffee.tex
Add elasticity plots
Rewrite partially the elasticity section

The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve. Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase). The amount of time spent in assembly and solve varies from problem to problem. As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g. because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what impacts the solve are, for instance, polynomial order and mesh discretization). The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e. known functions that are prescribed with the finite element specification). We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$). More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).

The application runs on a tetrahedral mesh composed of 196608 cells. In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$. To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.

In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, without and with optimizations enabled. There are four groups of two bars, each group corresponding to a specific test case; for example, the name $f_1p_1$ below the first group of bars refers to the case in which the elasticity problem has 1 coefficient function and has polynomial order 1. A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), to solve the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions). The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.

We first notice that, in the scenario $f_1p_1$, the assembly is dominated by matrix insertion: despite the application of generalized loop-invariant code motion, data alignment and padding, there is only a minimal performance gain. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. The transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}: this was achieved by a combined application of generalized loop-invariant code motion, alignment and padding, and expression splitting. Interesting are two facts. First, generalized loop-invariant code motion was particularly intensive: ... . Second, ... .

%We conclude noting that in explicit finite element methods you would have more local assembly , more often, no solver

>>>>

The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve. Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase). The amount of time spent in assembly and solve varies from problem to problem. As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g. because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what instead impact the solve are, for instance, polynomial order and mesh discretization). The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e. additional known functions that characterize the equation). We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$). More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).

The application runs on a tetrahedral mesh composed of 196608 elements. In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$. To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.

\begin{figure}[t]
	\centering
	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p1.pdf}}
	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p2.pdf}}
	\caption{Performance improvement over non-optimized code for the static linear elasticity equation. The tetrahedral mesh is composed of 196608 elements. Experiments were executed on a single core of a Sandy Bridge architecture.}
	\label{fig:full-app-res}
\end{figure}

In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, split over two figures (one for each polynomial order), without and with optimizations enabled. A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), in solving the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions). The \emph{non-opt} and \emph{opt} bars correspond, respectively, to the cases in which no optimizations and a combination of optimizations have been applied. The optimizations applied are generalized loop-invariant code motion, alignment and padding, and expression splitting. The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, there is only a minimal performance gain. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$. However, it is notably important noting that, as we anticipated, full-application speed ups rise proportionally with the time spent in assembly and, therefore, with the complexity of the equation. Further, by increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.
